---
title: 'Blog 6: The Ground Game - Campaigns'
author: June Park
date: '2022-10-15'
slug: []
categories: []
tags: []
---

```{r libraries, include = FALSE}
# hide stuff
knitr::opts_chunk$set(echo = FALSE)

# libraries
library(tidyverse)
library(ggplot2)
library(blogdown)
library(stargazer)
library(readr)
library(lubridate)
library(rmapshaper)
library(sf)
library(janitor)
library(leaflet)
library(stringr)
library(ggthemes)
library(gridExtra)
```

```{r data, include = FALSE}
cvap <- read_csv("cvap_district_2012-2020_clean.csv")
electionresults <- read_csv("incumb_dist_1948-2020 (3).csv")
genericballot <- read_csv("polls_df.csv")
gdp_q <- read_csv("GDP_quarterly.csv")
incumbency <- read_csv("incumb.csv")
```

## Introduction
This week, I explore the effects of voter turnout on the Democratic vote share at the Congressional district level. Working together with Lucy Ding and Kaela Ellis, we created models for all 435 Congressional districts (for the first time this semester!). A more extensive look at all of our combined work can be found in this [presentation](https://docs.google.com/presentation/d/1BylNhNNvf7J2S74J9gEgFVtR0okc9q5-CjXD8OkvbeY/edit?usp=sharing), which we presented in class on Tuesday, October 18 to GOV 1347. 

The models with voter turnout contain election data from 2012-2022. While, the models without voter turnout are more extensive with data all the way back to 1950. 

```{r adding turnout and generic ballot, include = FALSE}
# renaming geoid to st_cd_fips for left join
cvap <- cvap %>% 
  rename(st_cd_fips = geoid) 

# create turnout dataframe by left joining election results and cvap
turnout <-left_join(electionresults, cvap, by = c("year", "st_cd_fips")) %>% 
  select(year, state.x, district_num, winner_party, 
         RepVotes, DemVotes, DemVotesMajorPercent, st_cd_fips, cvap) %>% 
  filter(year >= 2012) %>%  # filter for 2012-2020
  rename(state = state.x) %>% 
  mutate(turnout = (RepVotes + DemVotes) / cvap * 100) # turnout column

# creating polls_d dataframe by cleaning genericballots
genericballot$poll_date <- as.Date(genericballot$poll_date, format = "%m/%d/%Y")

cleaned_polls <- genericballot %>% 
  filter(year >= 2012,
         days_until_election <= 52) %>% 
         # the closest poll to election day for 1952 was 51.5
  group_by(year, party) %>% 
  summarise(average_support = mean(support))

polls_d <- cleaned_polls %>% 
  filter(party =="D")

# adding generic ballot average to turnout
turnout <- left_join(turnout, polls_d, by = "year") %>% 
  select(year, state, district_num, winner_party, RepVotes, DemVotes,
         DemVotesMajorPercent, st_cd_fips, cvap, turnout, average_support)
```

```{r adding economy and incumbency, include = FALSE}
# Filter out data for Q6
gdp_q6 <- gdp_q %>%
  filter(year >= 2012, 
         quarter_cycle == 6) %>% 
  rename(gdp_q6 = GDPC1)

# Filter out data for Q7
gdp_q7 <- gdp_q %>%
  filter(year >= 2012, 
         quarter_cycle == 7) %>% 
  rename(gdp_q7 = GDPC1)

# Creating gdp_q_final
gdp_q_final <- left_join(gdp_q7, gdp_q6, by = "year") %>% 
  mutate(gdp_q7_q6 = (gdp_q7 - gdp_q6) / gdp_q6 * 100) %>% # percent difference
  rename(gdp_percent_difference = gdp_q7_q6) %>% 
  select(year, gdp_percent_difference)

# Creating incumbency
incumbency <- electionresults %>%  
  mutate(incumb = ifelse(winner_candidate == "DemCandidate",1,0)) %>% 
  mutate(year = year + 2) %>% 
  select (year, st_cd_fips, incumb)

# merging gdp_q_final and turnout
turnout <- left_join(turnout, gdp_q_final, by = "year")

# merging incumbency and turnout
turnout <- left_join(turnout, incumbency, by = c("year", "st_cd_fips")) %>% 
  select(year, state, district_num, winner_party, RepVotes, DemVotes,
         DemVotesMajorPercent, st_cd_fips, cvap, turnout, 
         average_support, gdp_percent_difference, incumb)

state_dists <- unique(turnout$st_cd_fips)
```

```{r 2022 dataframe}
# creating 2022 dataframe
# first, add the average voter turnout in district from 2012-2020
turnout_2022 <- turnout %>% 
  group_by(st_cd_fips) %>% 
  mutate(historical_turnout_by_district = mean(turnout)) %>% 
  ungroup() %>% 
  select(year, state, district_num, st_cd_fips, historical_turnout_by_district) %>% 
  filter(year == 2012) %>% 
  select(-year) %>% 
  unique()

# second, add nationwide variables
# Generic ballot: 45.6% for Democrats from FiveThirtyEight
# GDP: -0.6 (taken from Q2 change)
turnout_2022 <- turnout_2022 %>% 
  mutate(average_support = 45.6,
         gdp_percent_difference = -0.6)

# Incumbency
incumbency_2022 <- incumbency %>% 
  filter(year == 2022)

# Final 2022 data set to predict on
turnout22 <- left_join(turnout_2022, incumbency_2022, by = "st_cd_fips") %>% 
  select(state, district_num, st_cd_fips, historical_turnout_by_district,
         average_support, gdp_percent_difference, incumb) %>% 
  unique()
``` 

```{r}
# from Lucy
turnout14 <- turnout %>%
  filter(year == 2014) %>%
  select(year, turnout, st_cd_fips) %>%
  distinct() %>%
  distinct(st_cd_fips, .keep_all = TRUE)
turnout18 <- turnout %>%
  filter(year == 2018) %>%
  select(year, turnout, st_cd_fips) %>%
  distinct() %>%
  distinct(st_cd_fips, .keep_all = TRUE)
turnout22_fixed <- left_join(turnout14, turnout18, by = "st_cd_fips") %>%
  mutate(hist_avg = (turnout.x + turnout.y)/2) %>%
  select(st_cd_fips, hist_avg)
turnout22 <- left_join(turnout22, turnout22_fixed, by = "st_cd_fips") %>%
  select(-historical_turnout_by_district)
```

```{r, warning = FALSE}
# Instantiate lists
models <- list()
r2 <- c()
preds <- c()
preds_high <- c()
preds_low <- c()
coeff_support <- c()
coeff_turnout <- c()
coeff_gdp <- c()
coeff_incumb <- c()

# Loop through each district, model and predict
for (sd in state_dists){
  # Filter for dataframe with just district sd
  temp <- turnout %>%
    filter(st_cd_fips == sd)
  
  # Fit linear model for dem two-party vote-share
  m <- lm(DemVotesMajorPercent ~ average_support + turnout + gdp_percent_difference + incumb, data = temp)
  
  # Add model to list of models
  models[[length(models)+1]] <- m
  
  # Find r^2, coeffs, and add to list
  r2 <- c(r2, summary(m)$adj.r.squared)
  coeff_support <- c(coeff_support, summary(m)$coefficients[2])
  coeff_turnout <- c(coeff_turnout, summary(m)$coefficients[3])
  coeff_gdp <- c(coeff_gdp, summary(m)$coefficients[4])
  coeff_incumb <- c(coeff_incumb, summary(m)$coefficients[5])
  
  # Make a 2022 prediction and append to list
  new_data <- turnout22 %>%
    filter(st_cd_fips == sd) %>%
    rename(turnout = hist_avg)
  preds <- c(preds, predict(m, new_data))
  preds_high <- c(preds_high, predict(m, new_data %>% mutate(turnout = turnout + 8)))
  preds_low <- c(preds_low, predict(m, new_data %>% mutate(turnout = turnout - 8)))
}

output <- tibble(state_dists, preds, preds_high, preds_low, r2, coeff_gdp, coeff_incumb, coeff_support, coeff_turnout)
```

```{r Wyoming example}
# Example linear regression model: Wyoming
stargazer(m, type = "text")
```

```{r turnout prediction df, include = FALSE}
preds_df <- tibble(state_dists, preds, preds_high, preds_low)
```

```{r}
# low turnout 
length(preds[preds_low <= 50]) # 215 Republican seats
length(preds[preds_low > 50]) # 220 Democratic seats

# average turnout
length(preds[preds <= 50]) # 221 Republican seats
length(preds[preds > 50]) # 214 Democratic seats

# average turnout
length(preds[preds_high <= 50]) # 224 Republican seats
length(preds[preds_high > 50]) # 211 Democratic seats
```

```{r}
# Actual 2018 House Election Results
ggplot(turnout %>% filter(year == 2018), aes(x = DemVotesMajorPercent)) +
  geom_histogram(bins = 30, color = "white") +
  theme_few() +
  labs(title = "Actual 2018 House Election Results",
       subtitle = "House Districts",
       x = "Democratic Two-Party Vote-Share (%)",
       y = "Count") +
  ylim(0,50)
```

```{r}
# 2022 Election Predictions - Turnout Model
# Low Turnout

ggplot(output %>% mutate(preds_low = ifelse(preds_low > 100, 100, preds_low),
                                        preds_low = ifelse(preds_low < 0, 0, preds_low)), 
       aes(x = preds_low)) +
  geom_histogram(bins = 30, color = "white") +
  theme_few() +
  labs(title = "2022 Election Predictions - Turnout Model",
       subtitle = "Low Turnout",
       x = "Democratic Two-Party Vote-Share (%)",
       y = "Count") +
  xlim(-5, 105) +
  ylim(0,50)
```

```{r}
# 2022 Election Predictions - Turnout Model
# Average Turnout

ggplot(output %>% mutate(preds = ifelse(preds > 100, 100, preds),
                                        preds = ifelse(preds < 0, 0, preds)), 
       aes(x = preds)) +
  geom_histogram(bins = 30, color = "white") +
  theme_few() +
  labs(title = "2022 Election Predictions - Turnout Model",
       subtitle = "Average Turnout",
       x = "Democratic Two-Party Vote-Share (%)",
       y = "Count") +
  xlim(-5, 105) +
  ylim(0,50)
```

```{r}
# 2022 Election Predictions - Turnout Model
# High Turnout

ggplot(output %>% mutate(preds_high = ifelse(preds_high > 100, 100, preds_high),
                                        preds_high = ifelse(preds_high < 0, 0, preds_high)), 
       aes(x = preds_high)) +
  geom_histogram(bins = 30, color = "white") +
  theme_few() +
  labs(title = "2022 Election Predictions - Turnout Model",
       subtitle = "High Turnout",
       x = "Democratic Two-Party Vote-Share (%)",
       y = "Count") +
  xlim(-5, 105) +
  ylim(0,50)
```

```{r}
# R-Squared Values for Turnout Models in Each District

ggplot(output, aes(x = r2)) +
  geom_histogram(bins = 30, color = "white") +
  theme_few() +
  labs(title = "R-Squared Values for Turnout Models in Each District",
       subtitle = "House Districts",
       x = "R-Squared",
       y = "Count")
```

```{rm include = FALSE}
# coefficient models
p1 <- ggplot(output, aes(x = coeff_gdp)) +
  geom_histogram(bins = 30) + 
  theme_few() +
  labs(x = "GDP Coefficient",
       y = "Count")
p2 <- ggplot(output, aes(x = coeff_support)) +
  geom_histogram(bins = 30) +
  theme_few() +
  labs(x = "Generic Ballot Coefficient",
       y = "Count")
p3 <- ggplot(output, aes(x = coeff_turnout)) +
  geom_histogram(bins = 30) + 
  theme_few() +
  labs(x = "Turnout Coefficient",
       y = "Count")
grid.arrange(p1, p2, p3)
```

```{r}
p1 <- ggplot(output, aes(x = coeff_gdp)) +
  geom_histogram(bins = 30, color = "white") + xlim(-2.5, 2.5) + 
  geom_vline(xintercept = median(output$coeff_gdp, na.rm = TRUE),
             color = "red", size = 1) + 
  theme_few() +
  labs(x = "GDP Coefficient",
       y = "Count") +
  ylim(0, 130)

p2 <- ggplot(output, aes(x = coeff_support)) +
  geom_histogram(bins = 30, color = "white") + xlim(-10, 10) +  
  geom_vline(xintercept = median(output$coeff_support, na.rm = TRUE),
             color = "red", size = 1) + 
  theme_few() +
  labs(x = "Generic Ballot Coefficient",
       y = "Count") +
  ylim(0, 130)

p3 <- ggplot(output, aes(x = coeff_turnout)) +
  geom_histogram(bins = 30, color = "white") + xlim(-2.5, 2.5) + 
  geom_vline(xintercept = median(output$coeff_turnout, na.rm = TRUE),
             color = "red", size = 1) + 
  theme_few() +
  labs(x = "Turnout Coefficient",
       y = "Count") +
  ylim(0, 130)

p4 <- ggplot(output, aes(x = coeff_incumb)) +
  geom_histogram(bins = 30, color = "white") + xlim(-20, 20) + 
  geom_vline(xintercept = median(output$coeff_incumb, na.rm = TRUE),
             color = "red", size = 1) + 
  theme_few() +
  labs(x = "Incumbency Coefficient",
       y = "Count") +
  ylim(0, 130)

grid.arrange(p1, p2, p3)
```

