---
title: Post Election Reflection
author: June Park
date: '2022-11-20'
slug: []
categories: []
tags: []
---

```{r libraries, include = FALSE}
# hide stuff
knitr::opts_chunk$set(echo = FALSE)

# libraries
library(tidyverse)
library(ggplot2)
library(blogdown)
library(stargazer)
library(readr)
library(lubridate)
library(rmapshaper)
library(sf)
library(janitor)
library(leaflet)
library(stringr)
library(ggthemes)
library(gridExtra)
library(plm)
library(usdata)
library(ggrepel)
library(usmap)
library(tigris)
library(boot)
library(Metrics)
```

```{r data, include = FALSE}
# read in data
headers <- read_csv("2022_4_0_3.csv", col_names = FALSE, n_max = 1)
dat <- read_csv("2022_4_0_3.csv", skip = 2, col_names = FALSE)
colnames(dat) <- headers
district_names <- read_csv("district_names.csv")
```

```{r from enos, include = FALSE}
## hand enter MA, ME, and MS
dat[dat$FIPS == 25901,'Democratic'] = 153081; dat[dat$FIPS == 25901,'Republican'] = 96415
dat[dat$FIPS == 25902,'Democratic'] = 177519; dat[dat$FIPS == 25902,'Republican'] = 90791
dat[dat$FIPS == 25903,'Democratic'] = 145379; dat[dat$FIPS == 25903,'Republican'] = 82568
dat[dat$FIPS == 25904,'Democratic'] = 0; dat[dat$FIPS == 25904,'Republican'] = 0
dat[dat$FIPS == 25905,'Democratic'] = 198126; dat[dat$FIPS == 25905,'Republican'] = 70570
dat[dat$FIPS == 25906,'Democratic'] = 189753; dat[dat$FIPS == 25906,'Republican'] = 107384
dat[dat$FIPS == 25907,'Democratic'] = 144902; dat[dat$FIPS == 25907,'Republican'] = 26481
dat[dat$FIPS == 25908,'Democratic'] = 184076; dat[dat$FIPS == 25908,'Republican'] = 80961
dat[dat$FIPS == 25909,'Democratic'] = 192941; dat[dat$FIPS == 25909,'Republican'] = 131774

dat[dat$FIPS == 23901,'Democratic'] = 218630; dat[dat$FIPS == 23901,'Republican'] = 128996
dat[dat$FIPS == 23902,'Democratic'] = 151440; dat[dat$FIPS == 23902,'Republican'] = 140895

dat[dat$FIPS == 28901,'Democratic'] = 44387; dat[dat$FIPS == 28901,'Republican'] = 119673
dat[dat$FIPS == 28902,'Democratic'] = 107071; dat[dat$FIPS == 28902,'Republican'] = 71380
dat[dat$FIPS == 28903,'Democratic'] = 51628; dat[dat$FIPS == 28903,'Republican'] = 126081
dat[dat$FIPS == 28904,'Democratic'] = 39292; dat[dat$FIPS == 28904,'Republican'] = 122128

## LA and FL races with no reporting b/c no contest
dat[dat$FIPS == 22904,'Democratic'] = 0; dat[dat$FIPS == 22904,'Republican'] = 0
dat[dat$FIPS == 12905,'Democratic'] = 0; dat[dat$FIPS == 12905,'Republican'] = 0

# hard coded results for Utah's 3rd district
dat[dat$FIPS == 49903,'Democratic'] = 82865; dat[dat$FIPS == 49903,'Republican'] = 181141

# 3 races do NOT have any election information in this data set
  # 1205 : Florida 5th
  # 2204 : Louisiana 4th
  # 2504 : Massachusetts 4th
# reflectiondf %>% 
  # filter(is.na(DemVotesMajorPercent))
```

```{r cleaning data, include = FALSE}
# add leading 0 to create st_cd_fips
dat <- dat %>% 
  clean_names() %>% 
  mutate(cd = ifelse(geographic_name == "At Large", "00", cd)) %>%
  mutate(st_cd_fips = paste0(str_pad(state_fips, 2, pad = "0"), cd))

# selecting columns
# creating DemVotesMajorPercent
dat <- dat %>% 
  select(st_cd_fips, geographic_name, total_vote, rankd, rankr,
         democratic, republican, ca_democratic, ca_republican) %>% 
  mutate(democratic_all = democratic + ca_democratic,
         republican_all = republican + ca_republican, 
         twopartyvote = democratic_all + republican_all,
         DemVotesMajorPercent = democratic_all / twopartyvote * 100)

# hard code DemVotesMajorPercent for 3 districts (that originally had NAs in DemVotesMajorPercent)
# Florida 5th
dat[dat$st_cd_fips == 1205,'DemVotesMajorPercent'] = 0

# Louisiana 4th
dat[dat$st_cd_fips == 2204,'DemVotesMajorPercent'] = 0

# Massachusetts 4th
dat[dat$st_cd_fips == 2504,'DemVotesMajorPercent'] = 100
dat[dat$st_cd_fips == 2504,'rankd'] = 1 # to indicate the Democrat won

```

## Introduction
This week, I reflect on my predictive model for the 2022 House of Representatives elections based on the actual results of the 2022 elections. To summarize, I predicted that the Democrats would win 216 seats, while the Republicans would win 219 seats, resulting in a slim Republican majority in the House of Representatives. After reflecting on and analyzing my model based on the actual results, I find that my nationwide results (216 vs 219 seats) is actually quite close to the actual results (213 vs 219; 4 seats uncalled as of November 22, 2022), but my district-by-district results were not so accurate.

```{r adding my predictions, include = FALSE}
# reading and cleaning in my 2022 predictions
preds_2022 <- read_csv("preds_2022.csv")
preds_2022 <- preds_2022 %>% 
  rename(st_cd_fips = state_dists)

# combining actual results and my predictions
reflectiondf <- left_join(dat, preds_2022, by = "st_cd_fips")

# filter out 8 districts that somehow weren't calculated in my model LOL
# manually compute their numbers later using my model equation
reflectiondf_cleaned <- reflectiondf %>% 
  filter(!is.na(preds))
```

## Recapping my 2022 prediction model
For a more in-depth description and discussion about my model, please visit [my last blog post](https://junekimpark.github.io/election-blog/post/2022-11-04-prediction-for-the-2022-house-midterms/). 

Here I will present a short summary of my model: I ran a pooled linear regression model for each of the 435 districts to predict the Democratic two-party vote share at the district level. Then, I aggregated the data so that any districts with less than or equal to 50% for Democratic two-party vote share would be considered a Republican seat, while any districts with more than 50% for Democratic two-party vote share would be considered a Democratic seat. This allowed me to predict the Democratic Party’s seat share at the national level.

My independent variables were:
1. Generic Ballot
2. GDP % Difference
3. Dem Incumbent in District
4. Dem Party is in Power in House
5. Dem President is in Power
6. Previous Dem Two Party Vote Share
7. Expert Ratings

Using my model, I predicted that the **Democrats would win 216 seats, while the Republicans would win 219 seats**, resulting in a slim Republican majority in the House of Representatives. 

My **lower bound** at the 95% confidence level predicted that Democrats would win 213 seats, while the Republicans would win 222 seats, resulting in a larger Republican victory than my prediction. 

My **upper bound** predicted that Democrats would win 218 seats, while the Republicans would win 217 seats, resulting in the slimmest Democratic victory possible. 

```{r}
# Root Mean Squared Error = 14.85908
rmse(reflectiondf_cleaned$DemVotesMajorPercent, reflectiondf_cleaned$preds)

# HOW MANY DISTRICTS DID I PREDICT WRONG
wrong_districts <- reflectiondf_cleaned %>% 
  mutate(DemWinnerActual = ifelse(DemVotesMajorPercent > 50, 1, 0),
         DemWinnerPredicted = ifelse(preds > 50, 1, 0),
         MissedDistricts = ifelse(DemWinnerActual != DemWinnerPredicted, 1, 0)) %>% 
  select(st_cd_fips,
         DemVotesMajorPercent,
         preds,
         DemWinnerActual,
         DemWinnerPredicted,
         MissedDistricts) %>% 
  filter(MissedDistricts == 1)

left_join(wrong_districts, district_names, by = "st_cd_fips") %>% 
  select(state, district_num, DemVotesMajorPercent, preds) %>% 
  rename(State = state,
         District = district_num,
         "Actual Dem Vote Share" = DemVotesMajorPercent,
         "Predicted Dem Vote Share" = preds)
```

